---
title: "OrthoWatch — post-market surveillance of orthopedic device reports"
subtitle: "FDA MAUDE, 2020–2024 · control-chart trending · PRR/ROR signals · narrative mining"
date: today
format:
  html:
    embed-resources: true     # one self-contained file: charts, styles, all
    toc: true
    toc-title: "Sections"
    theme: cosmo
execute:
  echo: false                 # a report shows results; the code lives in R/
  warning: false
  message: false
---

```{r setup}
# The report is an EXECUTABLE DOCUMENT: every number and figure below
# is computed from the database at render time — nothing pasted, so
# the report can never drift out of date from the data.
# Rendered from report/, the project root is one level up:
knitr::opts_knit$set(root.dir = "..")
```

```{r load}
suppressPackageStartupMessages({
  library(DBI); library(RSQLite)
  library(dplyr); library(tibble); library(tidyr); library(stringr)
  library(ggplot2); library(knitr)
})
source("R/trending.R")
source("R/signal_detection.R")
source("R/text_mining.R")
source("R/run_history.R")

con <- dbConnect(RSQLite::SQLite(), "data/processed/orthowatch.db")

monthly <- read_result(con, "monthly_trends") |>
  as_tibble() |>
  mutate(month_date = as.Date(paste0(year_month, "-01")))
signals <- read_result(con, "signal_stats") |>
  as_tibble() |>
  mutate(evans_signal = as.logical(evans_signal))
terms <- read_result(con, "narrative_terms") |> as_tibble()
fam_counts <- dbGetQuery(con,
  "SELECT device_family, COUNT(*) AS n FROM clean_events
   GROUP BY device_family ORDER BY n DESC") |> as_tibble()
n_total <- sum(fam_counts$n)
date_range <- range(monthly$year_month)
dbDisconnect(con)
```

## What this is

`r format(n_total, big.mark = ",")` cleaned adverse event reports on
orthopedic devices, drawn from the FDA's MAUDE database
(`r date_range[1]` to `r date_range[2]`) and screened three ways: *when*
did reporting move (control charts), *what* problems attach to *which*
device families (disproportionality statistics), and what the
free-text narratives say that the coded fields don't (text mining).

**Read this first.** MAUDE reports are unverified reporter
submissions. Counts reflect reporting behavior as much as device
performance; report volume says nothing about how many devices are in
use. Every flag and signal below means *investigate* — never
*unsafe* — and nothing in this report is a finding about any product.

```{r fam-table}
fam_counts |>
  mutate(n = format(n, big.mark = ",")) |>
  kable(col.names = c("Device family", "Reports"),
        caption = "Cleaned reports per device family.")
```

## When: monthly trends against control limits

Each family's monthly report count, with a band at the mean ± 3
standard deviations — the century-old control-chart convention:
ordinary randomness lives inside the band; points outside deserve a
question.

```{r trends-plot, fig.width=9, fig.height=10}
plot_trends(monthly)
```

```{r flags-table}
monthly |>
  filter(status != "within limits") |>
  mutate(sd_out = round((n - center) / sigma, 1)) |>
  arrange(desc(abs(sd_out))) |>
  head(10) |>
  select(device_family, year_month, n, center, status, sd_out) |>
  kable(col.names = c("Family", "Month", "Reports", "Average",
                      "Status", "SD from avg"),
        caption = "The ten most extreme flagged months.")
```

Two caveats the numbers themselves taught: these series **trend**
strongly, and a single five-year mean over-flags trending series
(`r sum(monthly$status != "within limits")` of `r nrow(monthly)`
months flag — rolling-baseline limits are the documented refinement
on the roadmap); and the tallest towers here turned out, on
investigation, to reflect **reporting behavior** — batch submissions
concentrated in single reporters — rather than devices deteriorating.

## What: disproportionality signals (PRR/ROR)

For every (family, problem) pair, the 2×2-table statistics standard in
vigilance, judged by the published Evans rule (PRR ≥ 2, χ² ≥ 4, at
least 3 reports). Because each family is its own denominator, these
are immune to the volume trends above.

```{r signals-plot, fig.width=9, fig.height=10}
plot_top_signals(signals)
```

```{r signals-table}
signals |>
  filter(evans_signal) |>
  arrange(desc(prr)) |>
  head(10) |>
  mutate(across(c(prr, chi2, ror, ror_lo, ror_hi), ~ round(.x, 1))) |>
  select(device_family, product_problem, a, prr, ror, ror_lo, ror_hi) |>
  kable(col.names = c("Family", "Problem", "Reports (a)", "PRR",
                      "ROR", "CI low", "CI high"),
        caption = "The ten strongest signals by PRR.")
```

The `r sum(signals$evans_signal)` passing pairs cohere clinically —
a metal-degradation cluster for hips, slippage and integrity for
spinal fixation, wear and instability for knees, an
intraoperative-fit cluster for bone plates — and the method validates
itself: the dataset's largest, vaguest problem category signals for
no family, exactly as a category present everywhere should.
Over-reported ≠ over-occurring: signals open investigations, never
close them.

## What the stories say: narrative vocabulary

The coded fields are the tip of each report; the free-text narrative
is the substance. After tokenizing the narratives and stripping
filler, each family's most distinctive words (rate in the family vs.
everyone else — the same 2×2 arithmetic, applied to vocabulary):

```{r terms-plot, fig.width=10, fig.height=7}
plot_distinctive_terms(terms |>
  mutate(word_total = ave(n, word, FUN = sum),
         rate_family = per_10k / 1e4))
```

Text mining's two contributions here: it exposed a **subpopulation
the codes hid** (the bone-plate family contains a large
craniofacial/mandibular plating subgroup), and it **corroborated the
coded signals from an independent direction** (hip narratives orbit a
metal-on-metal resurfacing device class by name). It also settled two
open questions by reading: the vague-coded reports' stories often
contain specifics the codes omit, and one odd spinal signal traced to
a single reporter's legal-disclaimer template — a reporting pattern,
not a device pattern.

## How this report was made

This is an executable document: rendering it re-runs every query and
figure against `data/processed/orthowatch.db`, which is itself
rebuilt by a one-command pipeline (`Rscript run_pipeline.R`) whose
final stage is the project's `r 50`-test suite. Same raw files in,
same report out. Source, tutorials, and the full build story:
[github.com/akannan2987/orthowatch](https://github.com/akannan2987/orthowatch).
